Learning Excursion Sets of Vector-valued Gaussian Random Fields for Autonomous Ocean Sampling

### 2021-01-12 main_web.tex
Updates from Overleaf

The work presented here is primarily inspired by a case study
pertaining to using an AUV for spatial characterization of a frontal
system generated by a river plume. Fig.~\ref{fig:nidelven} shows the
survey area in Trondheim, Norway, where cold freshwater enters the fjord from a river, creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coriolis force the
cold fresh water tends to flow east. Depending on the variations in
river discharge, tidal effects, coastal current and wind, this
boundary often gets distorted, and knowledge about its location is
highly uncertain, making deterministic planning challenging. The goal
is therefore to use AUV measurements for improved description of the
interface between fresh and oceanic waters.  It is often not possible
to sample the biological variables of fundamental interest in such AUV
operations, but off-the-shelf instruments provide temperature and
salinity measurements which serve as proxies for the underlying
biological phenomenon. With the help of a vector-valued random field
model for temperature and salinity, one can then aim to describe the
plume.  The goal of plume characterization, in this way, relates to
that of estimating some regions of the domain, typically excursion
sets (ESs), when implicitly defined by the vector-valued random field.
In our context of environmental sampling, the joint salinity and
temperature excursions of a river plume help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}. Motivating examples for ESs of
multivariate processes are also abundant in other contexts, for
instance in medicine, where physicians do not rely solely on a single
symptom but must see several combined effects before making a
diagnosis.

@@

Other statistical work in the oceanographic domain include
\cite{wikle2013modern} focusing on hierarchical statistical models,
\cite{sahu2008space} studying spatio-temporal models for sea surface
temperature and salinity data and \cite{mellucci2018oceanic} looking
at the statistical prediction of features using an underwater glider.
In this work the main focus is not on statistical modeling per se, but
rather on statistical principles and computation underlying efficient
data collection. We combine novel possibilities in marine robotics
with spatial statistics and experimental design to provide useful AUV
sampling designs.

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Smith2016,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\citep{bueso1998state,zidek2019monitoring}. However, response variance and entropy being depending only in GRF models on measurement locations and not on response values, criteria based on them only tend to have limited flexibility for
active adaptation of trajectories based on measurement values.  The use
of data-driven adaptive criteria was introduced to include more
targeted sampling of regions of scientific interest in \cite{Low2009}
and \cite{fossuminformation}.

@@

Letting $\gp[\spatloc,\ell]$ denote the $\ell\text{-th}$ component of
$\gp[\spatloc]$ ($1\leq \ell\leq \no$), we use the term
\textit{generalized location} for the couple $x=(\spatloc,\ell)$.
%Given such a generalized location notation $x$,  
The notation $\gp[x]$ will be used to denote $\gp[\spatloc,\ell]$ 
and
%
will allow us to think of $\gp$ as a scalar-valued random field indexed by $\domain \times \{1\dots,p\}$, which will give the co-Kriging equations a particularly simple form that parallels the one of univariate Kriging. 
The letters $\spatloc$ and $\ell$ will be used for spatial locations and response indices respectively.
%
Furthemore, boldface letters will be used to denote concatenated
quantities corresponding to batches of observations.  Given a dataset
consisting of
$q$ observations at spatial locations
$\bm{\spatloc}=(\spatloc_1,\dots,\spatloc_q) \in
\domain^q$ and response indices $\bm{\ell}=(\ell_1,\dots, \ell_q)\in
\lbrace 1, ..., \no\rbrace^q$, %$i=1, ..., q$,
we use the concatenated notation
\begin{align*}
\bm{x}:=%(\bm{\spatloc}, \bm{\ell}):= 
(x_1,\dots, x_q),~\text{with }x_i=(\spatloc_i,\ell_i).
\end{align*}
We also compactly denote the field values at those different locations by
\begin{align*}
\gp[\bm{x}]:=
\left(\gp[\spatloc_1,\ell_1], ...,
\gp[\spatloc_q,\ell_{q}]\right) \in \mathbb{R}^{q}.
\end{align*}

@@

Assuming that $n$ batches of observations are available with sizes
$q_1,\dots, q_n$, and that one wishes to predict $\gp[\bm{x}]$ for
some batch of $q\geq 1$ generalized locations
$\bm{x}$, the simple co-Kriging mean then amounts to Kriging with respect to a
scalar-valued GRF indexed by $\domain\times \{1\dots,p\}$:
%
\begin{equation}\label{eq:cokrig_mean}
\mu_{[n]}(\bm{x})=\mu(\bm{x})+\lambda_{[n]}(\bm{x})^T (\mathbf{z}_{[n]}-\mu(\bm{x})).
\end{equation}
Here, $\mathbf{z}_{[n]}$ stands for the ($\sum_{i=1}^n
q_i$)-dimensional vector of observed (noisy) responses of
$Z$ at all considered generalized locations, and
$\lambda_{[n]}(\bm{x})$ is a vector of weights equal to
$$\left(k(\bm{x}_{[n]}, \bm{x}_{[n]})+\Delta_{[n]} \right)^{-1} k(\bm{x}_{[n]}, \bm{x})
$$
with $\bm{x}_{[n]}=(\bm{x}_1,\dots,
\bm{x}_n)$ and where
$\Delta_{[n]}$ is the covariance matrix of Gaussian-distributed noise
assumed to have affected measurements up to batch
$n$. For our applications with salinity and temperature observations,
this matrix is diagonal because we assume conditionally independent
sensor readings, but it might not be diagonal with other types of
combined measurements.  The matrix in parenthesis will be assumed to
be non-singular throughout the presentation. The associated
co-Kriging %(conditional)
residual (cross-)covariance function can also be expressed in the same
vein via
%
\begin{equation}\label{eq:cokrig_cov}
k_{[n]}(\bm{x},\bm{x}')=k(\bm{x},\bm{x}')-\lambda_{[n]}(\bm{x})^T 
\left(k(\bm{x}_{[n]}, \bm{x}_{[n]})+\Delta_{[n]} \right)
\lambda_{{[n]}}(\bm{x}').
\end{equation}

Let us now consider the case where a co-Kriging prediction of $Z$ was
made with respect to $n$ batches of generalized locations, concatenated again within $\bm{x}_{[n]}=(\bm{x}_1,\dots, \bm{x}_n)$,
and one wishes to update the prediction by incorporating a new vector
of observations $\mathbf{z}_{n+1}$ measured at a batch of
$q_{n+1} \geq 1$ generalized locations $\bm{x}_{n+1}$.
Thanks to our representation of co-Kriging in terms of simple Kriging
with respect to generalized locations, a strightforward adaptation of
the batch-sequential Kriging update formulae from
\citep{Chevalier.etal2013a} suggests that
% 
\begin{equation}\label{eq:meanCoK}
\mu_{[n+1]}(\bm{x})=\mu_{[n]}(\bm{x})+\lambda_{[n+1,n+1]}(\bm{x})^T (\mathbf{z}_{n+1}-\mu(\bm{x}_{n+1})),
\end{equation}
where $\lambda_{[n+1,n+1]}(\bm{x})$ denotes the $q_{n+1}$-dimensional
sub-vector extracted from $\lambda_{[n+1]}(\bm{x})$ that corresponds
to the Kriging weights for the last $q_{n+1}$ responses
when predicting at $\bm{x}$ relying on all measurements until batch $(n+1)$.
The associated co-Kriging residual (cross-)covariance function is
\begin{align}\label{eq:varCoK}
k_{[n+1]}(\bm{x},\bm{x}') & = k_{[n]}(\bm{x},\bm{x}')\\
 \nonumber - & \lambda_{[n+1,n+1]}(\bm{x})^T 
\left(k_{[n]}(\bm{x}_{n+1}, \bm{x}_{n+1})+\Delta_{n+1}\right)
%k_{[n]}(\bm{x}_{[n]}, \bm{x}_{[n]})
\lambda_{{[n+1,n+1]}}(\bm{x}'),
\end{align}
%
As noted in \citep{Chevalier2015} in the case of scalar-valued fields,
these update formulae naturally extend to universal Kriging in
second-order settings and apply without Gaussian assumptions. We will
now see how the latter formulae are instrumental in deriving
semi-analytical formulae for step-wise uncertainty reduction criteria
for vector-valued random fields.

@@

The case study considers the upper water column in the river plume,
represented by a two dimensional grid. Extensions to three-dimensional
domains are not methodologically different, but likely approximate
calculations by concentrating numerical integration on terms in the
vicinity of the autonomous vehicle \citep{fossum18b}. While we did not
consider any temporal effects, which would be relevant on a larger
time scale, we do consider the extension to spatio-temporal modeling
and envision that advection-diffusion equations could be useful
\citep{sigrist2015stochastic,richardson2017sparsity}. The sampling
capabilities are restricted by the sensors onboard the AUV, the
computational load, and navigation resources. The sampling of
temperature and salinity variables can be much higher than the grid
resolution, but, due to computational and navigation resources
available on the AUV, assimilation into a discretized fixed grid is
necessary. For more complex oceanographic phenomena, the methods will
need to be extended to non-Gaussian phenomena, possibly feature-based
mixtures of Gaussian processes which could potentially be run onboard
augmented by dynamical models. Running numerical models onboard a
robotic vehicle is currently infeasible, but high-resolution ocean
models or remote sensing data can be used to fit a more complex
statistical model \cite{davidson19}.

>>>>

The work presented here is primarily inspired by a case study
pertaining to using an AUV for spatial characterization of a frontal
system generated by a river plume. Fig.~\ref{fig:nidelven} shows the
survey area in Trondheim, Norway, where cold freshwater enters the fjord from a river, creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coriolis force the
cold fresh water tends to flow east. Depending on the variations in
river discharge, tidal effects, coastal current and wind, this
boundary often gets distorted, and knowledge about its location is
highly uncertain, making deterministic planning challenging. The goal
is therefore to use AUV measurements for improved description of the
interface between fresh and oceanic waters, which in turn enables more direct sampling of biological dynamics. It is often not possible
to sample these biological variables directly, but off-the-shelf instruments provide temperature and
salinity measurements which serve as proxies for the underlying
biological phenomenon. With the help of a vector-valued random field
model for temperature and salinity, one can then aim to describe the
plume. The goal of plume characterization, in this way, relates to
estimating some regions of the domain, typically excursion
sets (ESs), when implicitly defined by the vector-valued random field. The goal of plume characterization will in this setting relate to estimating an excursion set (ES), implicitly defined over the vector-valued random field. 
In our context of environmental sampling, the joint salinity and
temperature excursions of a river plume help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}. Motivating examples for ESs of
multivariate processes are also abundant in other contexts, for
instance in medicine, where physicians do not rely solely on a single
symptom but must see several combined effects before making a
diagnosis.

@@

Other statistical work in the oceanographic domain include
\cite{wikle2013modern} focusing on hierarchical statistical models,
\cite{sahu2008space} studying spatio-temporal models for sea surface
temperature and salinity data and \cite{mellucci2018oceanic} looking
at the statistical prediction of features using an underwater glider.
In this work the main focus is not on statistical modeling per se, but
rather on statistical principles and computation underlying efficient
data collection. We combine marine robotics
with spatial statistics and experimental design to provide rewarding AUV
sampling designs yielding scientific relevant data.

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Smith2016,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\citep{bueso1998state,zidek2019monitoring}. However, in GRF models response variance and entropy only depend on measurement locations and not on response values. Such criteria tend to have limited flexibility for
active adaptation compared to trajectories based on measurement values. The use
of data-driven adaptive criteria was introduced to include more
targeted sampling of regions of scientific interest in \cite{Low2009}
and \cite{fossuminformation}.

@@

Letting $\gp[\spatloc,\ell]$ denote the $\ell\text{-th}$ component of
$\gp[\spatloc]$ ($1\leq \ell\leq \no$), we use the term
\textit{generalized location} for the couple $x=(\spatloc,\ell)$.
%Given such a generalized location notation $x$,  
The notation $\gp[x]$ will be used to denote $\gp[\spatloc,\ell]$ 
and
%
will allow us to think of $\gp$ as a scalar-valued random field indexed by $\domain \times \{1\dots,p\}$, which will give the co-Kriging equations a particularly simple form that parallels the one of univariate Kriging. 
The letters $\spatloc$ and $\ell$ will be used for spatial locations and response indices respectively.
%
Furthermore, boldface letters will be used to denote concatenated
quantities corresponding to batches of observations.  Given a dataset
consisting of
$q$ observations at spatial locations
$\bm{\spatloc}=(\spatloc_1,\dots,\spatloc_q) \in
\domain^q$ and response indices $\bm{\ell}=(\ell_1,\dots, \ell_q)\in
\lbrace 1, ..., \no\rbrace^q$, %$i=1, ..., q$,
we use the concatenated notation
\begin{align*}
\bm{x}:=%(\bm{\spatloc}, \bm{\ell}):= 
(x_1,\dots, x_q),~\text{with }x_i=(\spatloc_i,\ell_i).
\end{align*}
We also compactly denote the field values at those different locations by
\begin{align*}
\gp[\bm{x}]:=
\left(\gp[\spatloc_1,\ell_1], ...,
\gp[\spatloc_q,\ell_{q}]\right) \in \mathbb{R}^{q}.
\end{align*}

@@

Assuming that $n$ batches of observations are available with sizes
$q_1,\dots, q_n$, and that one wishes to predict $\gp[\bm{x}]$ for
a batch of $q\geq 1$ generalized locations
$\bm{x}$, the simple co-Kriging mean then amounts to Kriging with respect to a
scalar-valued GRF indexed by $\domain\times \{1\dots,p\}$:
%
\begin{equation}\label{eq:cokrig_mean}
\mu_{[n]}(\bm{x})=\mu(\bm{x})+\lambda_{[n]}(\bm{x})^T (\mathbf{z}_{[n]}-\mu(\bm{x})).
\end{equation}
Here, $\mathbf{z}_{[n]}$ stands for the ($\sum_{i=1}^n
q_i$)-dimensional vector of observed (noisy) responses of
$Z$ at all considered generalized locations, and
$\lambda_{[n]}(\bm{x})$ is a vector of weights equal to
$$\left(k(\bm{x}_{[n]}, \bm{x}_{[n]})+\Delta_{[n]} \right)^{-1} k(\bm{x}_{[n]}, \bm{x})
$$
with $\bm{x}_{[n]}=(\bm{x}_1,\dots,
\bm{x}_n)$ and where
$\Delta_{[n]}$ is the covariance matrix of Gaussian-distributed noise
assumed to have affected measurements up to batch
$n$. For our applications with salinity and temperature observations,
this matrix is diagonal because we assume conditionally independent
sensor readings, but it might not be diagonal with other types of
combined measurements.  The matrix in parenthesis will be assumed to
be non-singular throughout the presentation. The associated
co-Kriging %(conditional)
residual (cross-)covariance function can also be expressed in the same
vein via
%
\begin{equation}\label{eq:cokrig_cov}
k_{[n]}(\bm{x},\bm{x}')=k(\bm{x},\bm{x}')-\lambda_{[n]}(\bm{x})^T 
\left(k(\bm{x}_{[n]}, \bm{x}_{[n]})+\Delta_{[n]} \right)
\lambda_{{[n]}}(\bm{x}').
\end{equation}

Let us now consider the case where a co-Kriging prediction of $Z$ was
made with respect to $n$ batches of generalized locations, concatenated again within $\bm{x}_{[n]}=(\bm{x}_1,\dots, \bm{x}_n)$,
and one wishes to update the prediction by incorporating a new vector
of observations $\mathbf{z}_{n+1}$ measured at a batch of
$q_{n+1} \geq 1$ generalized locations $\bm{x}_{n+1}$.
Thanks to our representation of co-Kriging in terms of simple Kriging
with respect to generalized locations, a strightforward adaptation of
the batch-sequential Kriging update formulae from
\citep{Chevalier.etal2013a} suggests that
% 
\begin{equation}\label{eq:meanCoK}
\mu_{[n+1]}(\bm{x})=\mu_{[n]}(\bm{x})+\lambda_{[n+1,n+1]}(\bm{x})^T (\mathbf{z}_{n+1}-\mu(\bm{x}_{n+1})),
\end{equation}
where $\lambda_{[n+1,n+1]}(\bm{x})$ denotes the $q_{n+1}$-dimensional
sub-vector extracted from $\lambda_{[n+1]}(\bm{x})$ that corresponds
to the Kriging weights for the last $q_{n+1}$ responses
when predicting at $\bm{x}$ relying on all measurements until batch $(n+1)$.
The associated co-Kriging residual (cross-)covariance function is
\begin{align}\label{eq:varCoK}
k_{[n+1]}(\bm{x},\bm{x}') & = k_{[n]}(\bm{x},\bm{x}')\\
 \nonumber - & \lambda_{[n+1,n+1]}(\bm{x})^T 
\left(k_{[n]}(\bm{x}_{n+1}, \bm{x}_{n+1})+\Delta_{n+1}\right)
%k_{[n]}(\bm{x}_{[n]}, \bm{x}_{[n]})
\lambda_{{[n+1,n+1]}}(\bm{x}'),
\end{align}
%
As noted in \citep{Chevalier2015} in the case of scalar-valued fields,
the update formulae naturally extends to universal Kriging in
second-order settings and apply without Gaussian assumptions. We will
now see how the latter formulae are instrumental in deriving
semi-analytical formulae for step-wise uncertainty reduction criteria
for vector-valued random fields.

@@

The case study considers the upper water column in the river plume,
represented by a two dimensional grid. Extensions to three-dimensional
domains are not methodologically different, but likely approximate
calculations by concentrating numerical integration on terms in the
vicinity of the autonomous vehicle \citep{fossum18b}. While we did not
consider any temporal effects, which would be relevant on a larger
time scale. We do consider the extension to spatio-temporal modeling
and envision that advection-diffusion equations could be a point of departure
\citep{sigrist2015stochastic,richardson2017sparsity}. The sampling
capabilities are restricted by the sensors onboard the AUV, the
computational load, and navigation resources. The sampling of
temperature and salinity variables can be much higher than the grid
resolution, but, due to computational and navigation resources
available on the AUV, assimilation into a discretized fixed grid is
necessary. For more complex oceanographic phenomena, the methods will
need to be extended to non-Gaussian phenomena, possibly feature-based
mixtures of Gaussian processes which could potentially be run onboard
augmented by dynamical models. Running numerical models onboard a
robotic vehicle is currently infeasible, but high-resolution ocean
models or remote sensing data can be used to fit a more complex
statistical model \cite{davidson19}.
